Incremental tactics

### 2012-06-08 doc/papers/LFMTP2012/paper.tex
paper: head -> tip + misc. little improvements

To write deltas by sharing subterms, we need to be able to address any
subterm and instantiate its free variables. For this, we define the
\emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
contextual metavariables $\meta X$, $\meta Y$\ldots\ as in
\cite{nanevski2008contextual}. Metavariables stand for well-typed
terms defined in a repository, and are instantiated wrt. their free
variables. A repository also contains a distinguished metavariable
corresponding to the head of the term stored in it. For instance, the
repository
$$
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

@@

We overload \emph{substitution} as a partial operation on objects,
given a parallel substitution $\sigma$. Since the syntax of object is
not stable by (textual) substitution, it is defined as
\emph{hereditary}\footnote{ Note that this notion of substitution,
  present in \cite{pfenning2007term} differs from
  \cite{hl07mechanizing} as it presupposes $\eta$-long normal forms:
  once a \emph{cut} starts, there must be exactly as many formal and
  actual arguments. A non-empty spine cut against another non-empty
  spine $\gcut {\app H S} {S'}$ blocks the substitution instead of
  reducing to $\app H {(\gconcat S S')}$ (an implicit
  $\eta$-expansion); conversely, exhausting actual but not formal
  arguments as in $\gcut{\lam x M}{\gnil}$ blocks the substitution
  too. For example, the reduction $\gcut {\lam x \app {\var x}{\var
      x}} {\lam x \app {\var x}{\var x}}$ gets stuck because the
  second $\var x$ in the left part is substituted by a function yet is
  not itself a function. }: it is mutually recursive with a \emph{cut}
function $\gcut M S$ which can in turn trigger a chain of
substitutions, leading to a canonical objects or an error.
\begin{align*}
  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
    \sigma)}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \app {\var x} {\gsubst S \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \gcut M {\gsubst S \sigma} &
  \text{if $\sigma(\var x) = M$}
  \\
  \gsubst {\smeta X {\sigma}} {\sigma'} &=
  \smeta X {\gcomp\sigma{\sigma'}}
  \\
  \gsubst\spinenil\sigma &=
  \spinenil \\
  \gsubst{\spinecons M S}\sigma &=
  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
  \\
  \gcut {\lam x M} {\spinecons N S} &=
  \gcut {\gsubst M {\msubst x N}} S
  \\
  % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
  % \gcut {\smeta X\sigma} S &=
  % \smeta X{\gcomp\sigma\sigma'}
  % \\
  \gcut {\app H S} \spinenil &=
  \app H S
  \\
  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
  \\
  \gcomp \sigma \msubstnil &=
  \sigma
  \\
  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
  \\
  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
  \\
  \gsubst {\msubstnil}{\sigma} &= \msubstnil
\end{align*}

>>>>

To write deltas by sharing subterms, we need to be able to address any
subterm and instantiate its free variables. For this, we define the
\emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
contextual metavariables $\meta X$, $\meta Y$\ldots\ as in
\cite{nanevski2008contextual}. Metavariables stand for well-typed
terms defined in a repository, and are instantiated wrt. their free
variables. A repository also contains a distinguished metavariable
corresponding to the ``tip'' of the term stored in it. For instance, the
repository
$$
\small
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

@@

We overload \emph{substitution} as a partial operation on objects,
given a parallel substitution $\sigma$. Since the syntax of object is
not stable by (textual) substitution, it is defined as
\emph{hereditary}\footnote{ Note that this notion of substitution,
  present in \cite{pfenning2007term}, differs from the more standard
  definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
  normal forms: once a \emph{cut} starts, there must be exactly as
  many formal and actual arguments. A non-empty spine cut against
  another non-empty spine $\gcut {\app H S} {S'}$ blocks the
  substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
  implicit $\eta$-expansion); conversely, exhausting actual but not
  formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
  substitution too. For example, the reduction $\gcut {\lam x \app
    {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
  because the second $\var x$ in the left part is substituted by a
  function yet is not itself a function. }: it is mutually recursive
with a \emph{cut} function $\gcut M S$ which can in turn trigger a
chain of substitutions, leading to a canonical objects or an error.
\begin{align*}
  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
    \sigma)}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \app {\var x} {\gsubst S \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \gcut M {\gsubst S \sigma} &
  \text{if $\sigma(\var x) = M$}
  \\
  \gsubst {\smeta X {\sigma}} {\sigma'} &=
  \smeta X {\gcomp\sigma{\sigma'}}
  \\
  \gsubst\spinenil\sigma &=
  \spinenil \\
  \gsubst{\spinecons M S}\sigma &=
  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
  \\
  \gcut {\lam x M} {\spinecons N S} &=
  \gcut {\gsubst M {\msubst x N}} S
  \\
  % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
  % \gcut {\smeta X\sigma} S &=
  % \smeta X{\gcomp\sigma\sigma'}
  % \\
  \gcut {\app H S} \spinenil &=
  \app H S
  \\
  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
  \\
  \gcomp \sigma \msubstnil &=
  \sigma
  \\
  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
  \\
  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
  \\
  \gsubst {\msubstnil}{\sigma} &= \msubstnil
\end{align*}
